---
title: "Use the canonical function, not your own"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use the canonical function, not your own}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This vignette exists because a drive-time access map was built for a midwifery
workforce study, and along the way its author re-implemented — badly — six
things that already existed in this ecosystem, tested. Each section below pairs
**what was written by hand** with **what should have been called**, so the next
person can skip the detour.

The packages, roughly outermost to innermost:

| Package | Holds |
|---|---|
| `mufflyaccess` | SSOT constants and safe arithmetic: bands, CONUS geography, denominators |
| `twostep` | E2SFCA accessibility, population-weighted coverage, access-language guard |
| `mysterymaps` | Leaflet map construction |
| `cliff` | Workforce cliff / retirement analysis |

The rule of thumb: **if a value feels like it belongs to the project rather than
to your script, it is probably already a constant somewhere.**

## 1. Geography constants

A map excluded non-contiguous states with a hand-typed FIPS vector:

```{r}
# hand-rolled
CONUS_EXCLUDE <- c("02", "15", "60", "66", "69", "72", "78")
counties <- dplyr::filter(counties, !STATEFP %in% CONUS_EXCLUDE)
```

That vector is byte-identical to the canonical one. It was retyped, not derived,
which means the next revision of either copy silently disagrees:

```{r}
library(mufflyaccess)

NON_CONTIGUOUS_FIPS
#> [1] "02" "15" "60" "66" "69" "72" "78"

counties <- dplyr::filter(counties, !STATEFP %in% NON_CONTIGUOUS_FIPS)
```

`CONUS_STATE_FIPS`, `CONUS_STATE_ABBR` and `NON_CONTIGUOUS_CODES` are there too.
For point data, `cliff::in_conus_bbox()` replaces a hand-written bounding box —
another thing that got retyped as `lat >= 17 & lat <= 72 & lon >= -180 & lon <= -64`.

## 2. Drive-time bands

```{r}
# hand-rolled
BANDS <- c(30, 60)
```

```{r}
CANONICAL_BANDS
#> [1]  30  60 120 180

get_canonical_bands()
get_primary_access_band()   # the band a headline result should quote
```

Using the constant makes an important thing visible: a map showing only 30 and
60 is showing *half* the canonical set. That is a defensible choice — the
midwifery study had no 120/180 polygons for newly-routed locations — but it
should be a stated deviation, not an invisible one.

## 3. Rates that do not blow up

Rate construction was written inline, three times, slightly differently:

```{r}
# hand-rolled
midwives_per_1k <- ifelse(!is.na(births) & births >= 50,
                          round(1000 * n_midwives / births, 2), NA_real_)
```

```{r}
midwives_per_1k <- safe_rate(n_midwives, births, multiplier = 1000, digits = 2)
```

`safe_divide()`, `safe_rate()`, `safe_ratio()`, `safe_percent()` all handle the
zero denominator, the `NA`, and the `Inf` that `round()` will happily carry into
a published table. `safe_divide(on_zero = "error")` turns a silent `NA` into a
loud failure where that is what you want.

## 4. Do not call it a shortage

`twostep` ships a vocabulary guard, and it encodes a real methodological
constraint: an accessibility surface without a defensible demand target cannot
support normative claims.

```{r}
library(twostep)

ACCESS_FORBIDDEN_TERMS
#> "shortage" "surplus" "adequacy" "adequate" "inadequate" "unmet need" ...

ACCESS_SAFE_LABELS
#> "modeled accessibility" "relative accessibility" "% of women with access" ...

assert_access_language(map_caption, context = "map caption")
```

Wire it into the render, not into a review checklist:

```{r}
caption <- "Ground within 60 minutes of a certified nurse-midwife"
assert_access_language(caption, context = "legend title")   # passes

bad <- "Counties with an inadequate midwifery supply"
assert_access_language(bad)                                  # stops
```

## 5. Coverage measured in people, not acres

A first draft reported coverage as **land area** — 1,759,430 km² within 30
minutes. Land area rewards empty terrain: a polygon over rangeland counts the
same as one over a city. The population-weighted version already exists:

```{r}
compute_band_tract_overlap(bands_sf, tracts_sf)
allocate_pop_areaweighted(tracts_sf, target_sf)
zero_access_share(access_tbl)
weighted_mean_all(x, w)
```

`TRACT_REACHED_COVERAGE_PCT` is the canonical threshold for calling a tract
reached, so two studies asking "what share of births are within 60 minutes?"
give comparable answers.

## 6. Rurality

```{r}
rurality_from_ruca(ruca_codes)
RUCA_NONMETRO_MIN
```

Written by hand, the metro/nonmetro cut is one `case_when()` and one chance to
put the boundary in the wrong place.

## 7. Coverage surfaces on a leaflet map

This is the one case where the canonical answer did **not** exist, and the
correct response was to add it here rather than keep a private copy.

The package's choropleth builders are organised around per-geography values.
A dissolved coverage union has none — it answers "is this ground covered?", not
"what is this county's number?" — so `mysterymaps_add_coverage_surfaces()` was
added:

```{r}
library(mysterymaps)
library(leaflet)

m <- leaflet() |>
  addProviderTiles("CartoDB.PositronNoLabels") |>
  addPolygons(data = counties, fillColor = ~pal(rate), group = "Supply") |>
  mysterymaps_register_base_legend("Supply", key = "supply") |>
  mysterymaps_add_coverage_surfaces(
    surfaces = list("Within 30 minutes" = union30,
                    "Within 60 minutes" = union60,
                    "Beyond 60 minutes" = gap60),
    colors        = c("#08519c", "#3182bd", "#d94801"),
    legend_labels = c("within 30 min", "within 60 min", "more than 60 min"),
    legend_titles = c("Drive-time coverage", "Drive-time coverage", "Coverage gap")) |>
  addLayersControl(baseGroups = c("Supply", "Within 30 minutes",
                                  "Within 60 minutes", "Beyond 60 minutes"))

m <- mysterymaps_base_legend_switcher(m, default = "Supply")
```

Two behaviours worth knowing:

**Base groups, not overlays.** Stacking a translucent coverage fill over a
viridis choropleth multiplies two colour scales into a third belonging to
neither. Coverage layers are alternative *views* of a map, not additions to it.

**`addLegend(group=)` does not work with base groups.** It follows overlay
groups only. A map with four base layers and four legends therefore renders all
four at once, stacked down the edge — a live bug in more than one map in this
ecosystem. `mysterymaps_base_legend_switcher()` attaches a `baselayerchange`
handler that shows only the active layer's legend.

## 8. Labels on a crowded point layer

The reflex for 11,792 points is marker clustering. Clustering replaces the data
with a count and makes the reader zoom repeatedly before learning anything:

```{r}
# what not to do
addCircleMarkers(..., clusterOptions = markerClusterOptions())
```

Draw every point on a canvas renderer and gate the *labels* instead:

```{r}
leaflet(options = leafletOptions(preferCanvas = TRUE)) |>
  addCircleMarkers(data = providers, label = ~full_name, group = "Providers",
                   options = pathOptions(pane = "pts")) |>
  mysterymaps_zoom_gated_labels("Providers", min_zoom = 9, max_labels = 400)
```

The distribution stays visible at every zoom; names appear only where they are
legible. Note `pathOptions(pane = "pts")` with `addMapPane("pts", zIndex = 650)`
— without the pane, markers draw above a choropleth but do not reliably
hit-test, so clicks land on the polygon underneath.

## 9. Zero is not a small number

```{r}
sc <- mm_jenks_zero_scale(counties$rate, k = 6, digits = 1)
addPolygons(fillColor = sc$color(counties$rate))
addLegend(colors = sc$leg_cols, labels = sc$leg_labs)
```

A hand-rolled scale binned `0.0–0.5`, so counties with **no** provider rendered
identically to counties with a low rate — 1,619 of 3,109 counties, over half the
map, in the class the study was about. `mm_jenks_zero_scale()` gives zero its
own colour and Jenks-classifies only the positive values.

## How to check before you write

```{r}
# Is there already a constant or helper for this?
ls("package:mufflyaccess")
ls("package:twostep")
ls("package:mysterymaps")

apropos("conus|band|rurality|safe_")
```

If a function seems missing, confirm it is genuinely missing rather than
differently named — and if it really is absent, add it to the package that
should own it. Keeping a private copy and documenting why your caller is special
is how this ecosystem ended up with three copies of a gender gate, where a fix
applied to one was a fix applied to none.
